---
title: "PS270 Final Project"
author: "Donovan Cerny"
description: "South Alabama First Ball Win Rate"
output:
  distill::distill_article:
    self_contained: false
---
```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
library(scales)
library(infer)
library(modelsummary)
```

## Introduction

Does serving to specific places on a volleyball court increase chances of winning the point on the first ball? In this study I plan to examine the extent to which serving to distinct places on the court produce a greater chance of winning the point on the first ball. I hypothesize that serving to specific spots on the court is a major factor in winning the first point. I expect that some zones, despite the player or the team, will be more likely than others to induce a win on the first ball. This is because in different rotations it is naturally harder to get a good pass for the setter to make an effective play allowing for the point to be won. Thus, serving in certain spots will be more effective for winning first points. My sample consists of scouting reports from 5 NCAA Division 1 women's volleyball teams. The explanatory variable in my data is the pass location which is coded as 1,2,3,4,5, and 6. Each number is the zone where the serve was passed from, which is where I make the assumption that where the serve was passed from is also where the ball was served to. My outcome variable is if the point was won or lost in the first point or the rally continued. 
   
If the observed data shows that the proportion of first points won is significantly greater for one of the serve locations than my hypothesis will be supported and I will reject the null. If all of the serve locations have fairly equal proportions of first points won, then I will have failed to find evidence for my hypothesis and will fail to reject the null. I find this to be particularly interesting because I used to play volleyball and always wondered if one zone was always more effective than others. This is also a very important aspect of the game because winning a point on the first ball limits the other team from going on big runs of scoring points and allows for your team to go on a run itself. Additionally, this project is real data that came from the scouting reports that my sisters team has provided (South Alabama) making the project extra interesting.

## Data Section

As stated before, the data used for this experiment comes from scouting reports that were shared with me from South Alabama university. Since the reports come from South Alabama, the data is in their point of view, so a serve to zone 1 that results in a first ball point win, means that South Alabama served to zone 1 and won the point on the first ball following the serve. Within these reports there is a lot of information that simply comes from watching what happens in live games and documenting it. This makes my research design cross sectional. I have decided to focus on the serving aspect of the reports. Unfortunately, there is no way to know exactly where the server was trying to serve the ball, so I have assumed the location the serve was passed from, was the intended target. The data also contains information on whether a team won or lost the point on the first ball. This is important because the first ball is the only point that is affected by the serve. Thus, by counting the number of total serves to a specific zone, then dividing that number by the number of times the first ball point was won after a serve to each specific zone, I get the proportion of first ball points won per zone. 
  
With the zone served as the predictor and the first ball point being won as the outcome, I can compare each zone to see if one zone has a higher percentage of first ball points won. If one zone's win percentage is clearly higher than the rest and I can prove that my observed outcome is significantly different from a randomized sample (which eliminates the relationship between zones and first ball wins by randomizing the data) then my hypothesis will be supportable. In order for my data to be statistically significant I will be judging based on a 95% confidence interval. If my observed data is not significantly different from the randomized data, then my hypotheses will be rejected because a small difference in the randomized data and my data would prove that the relationship between first ball wins and serve location is minimal if not obsolete.

### Organizing data:

```{r Loading data}
aub<- read_csv("Team_data/AUB_MASTER.csv") 
ccu<-read_csv("Team_data/CCU_MASTER.csv")
asu<-read_csv("Team_data/ASU_MASTER.csv")
gso<-read_csv("Team_data/GSO_MASTER.csv")
isu<-read_csv("Team_data/ISU_MASTER.csv")
```

```{r Selecting needed variables}
aub<- aub |> 
  select(Team, Rotation, `Pass Loc`, FBWL)
ccu<- ccu |> 
  select(Team, Rotation, `Pass Loc`, FBWL)
asu<- asu |> 
  select(Team, Rotation, `Pass Loc`, FBWL)
gso<- gso |> 
  select(Team, Rotation, `Pass Loc`, FBWL)
isu<- isu |> 
  select(Team, Rotation, `Pass Loc`, FBWL)
```

```{r Joining data}
all_teams<- bind_rows(aub, ccu, asu, gso, isu) # Combine all data sets into one tibble
```


```{r Removing NA Values}
all_teams<- all_teams |> 
   mutate(FBWL = case_when(FBWL == "W" ~ "Win", FBWL == "L" ~ "Loss"),
          FBWL = replace_na(FBWL, "Rally Continued")) |> 
  mutate(`Pass Loc` = paste("Zone", `Pass Loc`)) |>
  mutate(`Pass Loc` = ifelse(`Pass Loc` == "Zone NA", NA, `Pass Loc`)) |>
  filter(!is.na(`Pass Loc`))
```

### Plot of First Ball Wins:

```{r Plot of First Ball Wins}
all_teams|> 
  drop_na()|>
  group_by(`Pass Loc`, FBWL)|> 
  summarise(n = n(), .groups = "drop_last") |> 
  mutate(prop = n/sum(n)) |> 
  filter(FBWL == "Win" | FBWL == "Loss") |> 
  ggplot(aes(x = FBWL, y = n, fill = FBWL)) + 
  geom_col() + 
  scale_fill_manual(values = c("Win" = "dodgerblue1", "Loss" = "indianred")) +
  labs(title = "First Ball Wins Vs First Ball Losses", 
       x = "First Ball Wins and Losses", 
       y = "Number of Points", 
       fill = "Point Outcome")
```


## Results Section

### Win Proportions with Visualization:
 
```{r Proportion of wins for each serve location}
 prop_win<- all_teams|> 
  drop_na()|>
  group_by(`Pass Loc`, FBWL)|> 
  summarise(n = n(), .groups = "drop_last") |> 
  mutate(prop = n/sum(n)) |> 
  filter(FBWL == "Win")
```


```{r Plot for win proportion}
ggplot(prop_win, mapping = aes(x = `Pass Loc`, y = prop ))+ 
  geom_col(fill = "dodgerblue", alpha = 1) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Percentage of First Ball Points Won by Serve Location", 
       x = "Serve Location", 
       y = "Percent of First Ball Points Won")
```

### Regression:
```{r Regression showing zone win comparisons}
reg_win<- lm(FBWL == "Win" ~ `Pass Loc`, data = all_teams) 

fstat <- summary(reg_win)$fstatistic

model_p <- pf(fstat[1], fstat[2], fstat[3], lower.tail = FALSE)

modelsummary(reg_win,
  output = "markdown",
  title = "Regression Results for Serve Location",
  estimate = "{estimate}",
  statistic = "({p.value})",
  gof_map = data.frame(raw = c("nobs", "r.squared"),
  clean = c("Num.Obs.", "R2"),
  fmt = c(0, 3)),
  add_rows = data.frame(term = "Model p-value",
  estimate = formatC(model_p, format = "f", digits = 3)))
```

### Additional Analysis:

```{r Calculating the observed distribution value}
obv_dist <- all_teams |> 
  filter(FBWL == "Win"| FBWL == "Loss") |> 
  specify(FBWL ~ `Pass Loc`) |> 
  hypothesize(null = "independence") |> 
    calculate(stat = "Chisq")
```

```{r Creating and calculating null distribution values}
null_dist <- all_teams |> 
  filter(FBWL == "Win" | FBWL == "Loss") |> 
  specify(FBWL ~ `Pass Loc`) |> 
  hypothesise(null = "independence") |> 
  generate(reps = 1000, type = "permute") |> 
  calculate(stat = "chisq") 
```

```{r Calculating p-value for the null distribution}
null_p_value<- null_dist |> 
  get_p_value(obs_stat = obv_dist, direction = "right")
knitr::kable(null_p_value, col.names = "Null Distribution P-value")
```

```{r Graph null distribtion with p-value and observed value}
p_plot<- null_dist|> 
  visualise() + 
  shade_p_value(obs_stat = obv_dist, direction = "right", fill = "red", alpha = .5) +
  labs(title = "Null Distribution of Test Statistic",
       x = "Chi-Square Statistic",
       y = "Density")
p_plot
```

### Interpretation:

Looking at the plot of first ball win rates per zone, zone 2 and 4 both have higher win rates then the rest of the zones, with zone 2 being the highest. This is good for my hypothesis because it shows that maybe serving to zone 2 will provide an advantage and allow for a higher percentage of first ball points to be won. In order to solidify this observation of the graph I ran a regression which will tell me exactly how different the percentage of first ball wins is for each zone. However, this difference in win percentage could just be purely by chance, which is why I must also interpret the statistical significance of this observation. As previously mentioned, I will be doing this test with a 95% confidence interval meaning that the p-value of the data must be lower than 0.05 in order for this data to be statistically significant. If it is the case and the p-value is low enough, it means that this observation is so unlikely, that there is a chance for a causation factor between what zone is served and first ball wins.

Within this regression, the intercept is the predicted win rate for serves to zone 1, which is 0.302 or 30.2 and this serves as a baseline value for the other locations. This means that when the ball is served to zone 2, the predicted first ball win rate is .115 or 11.5 percentage points higher than if the ball were to be served to zone 1. Following this logic for the other zones, it can be observed that serves to zone 3 have a predicted win rate that is 11.7 percentage points lower than serves to zone 1. Serves to zone 4 have a predicted win rate that is 8.6 percentage points higher than serves to zone 1. Serves to zone 5 have a predicted win rate that is 0.25 percentage points lower than serves to zone 1. Finally, serves to zone 6 have a predicted win rate that is 0.9 percentage points lower than serves to zone 1. All in all, this regression predicts that serves to zone 2 will have a win rate that is marginally higher than zone 1 and the next closest predicted win rate, which is zone 4. However, without statistical significance, these differences may be due to random variation.

Moreover, the regression also shows the p-values for not only each zone, but also for the observation as a whole. The p-values for each zone vary but unfortunately every zone's p-value is over the 0.05 threshold. Additionally, the p-value for the observation as a whole is 0.326 meaning that my observation is way above 0.05 and thus, not statistically significant. All in all, this means I fail to reject the null hypothesis and fail to find evidence of causation for first ball wins and zone served to.

While my observation might have had a lack of evidence to reject the null, I thought I would try another test just in case I got a different result. For this test I choose to do a chi squared test, which would also allow me to establish causation due to its comparison with randomized data. This kind of text makes sense in this situation because it compares the chi squared value from the observation to many permutations of the data. These permutations take the data and randomize the results in the data which effectively removes any possible causation between serve location and first ball points won. This test allows me to see how far the observed value is from the expected value. If it is very far away and creates a p-value under 0.05 then I would have good evidence to reject the null, if it's not far from the expected value, then I will have a higher p-value and again have too little evidence to reject the null. Unfortunately, the p-value I got from this test was 0.36 which is still well above the 0.05 threshold for a 95% confidence interval meaning I still fail to reject the null and I do not have solid evidence to establish a causation between serve location and first ball point wins. This is also illustrated in a visual which shows the distribution of chi squared test scores, where the solid red line is my observed value. If my observed value would have been further to the right on this visual it would have illustrated that my observed value was so different from the average values of the randomized values that there must have been a causation.


## Conclusion

In the end, both my original regression on the observation and my chi squared test both failed to provide evidence of a serve location causing a greater proportion of first ball points to be won. This being said, I fail to reject my null hypothesis meaning in doing this test I found little evidence that serve location and first ball points being won are intertwined but not enough evidence to say the observation is statistically significant. Thus, the test shows that it is unlikely that serve location is causal of first ball points won. However, there were many limitations to my test. For starters, the whole test was based on observation and could not establish true causality even if it was statistically significant. It also worked off of the assumption that the serve was meant to be served to the zone it was passed from, and this is simply not always the case. There is also some missing data because this test was based on observations, some zones in my data were served significantly less times than others which could have caused some bias in the overall data. Furthermore, the test does not account for possible player and overall team disparities. For example, if one player or team is just much worse at passing and is constantly changing what zone they pass from, the data could be different. This is also an example of a confounder in this test because if a player is much worse at passing they will likely be targeted no matter which zone they are in and they will likely result in a loss. This means that one player could not only affect which zone is being served, but also if the team wins the first ball point. Given more time and money, I would treat this more like an experiment rather than an observation. Meaning I could tell the server where to serve and I could also ensure that each zone receives the same amount of serves to make the data even less biased. The analysis could also be improved by adding in all the different rotations, positions, and side of the court the serve is initiated from in order to get a more exact answer and remove a lot of unknown data from my analysis.




 
 
 
 
 
 
 